---
title: "Homework on Dynamic choice"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
---

The goal of this homework is to familiarize our-self with dynamic choice models. We are going to consider a simple case where agent choose to work or to not work. 

In this simple model we are going to consider the effects of simple policies.

# The model

Individuals will differ in their producitivity that will denote $x$. We will use a discrete support for it. Each period will choose wether to work or not to work. Work will give payment $w = x \cdot a$ where $a$ is an aggregate state which my move stochastically.

Finally transferring from unemployment to employment will only happen with a probability $\lambda$. 

Our value function is then indexed by $a$, $x$  and the current working status $e$.

Finally, we will consider an earning tax $\tau$ and an unemployment benefit $b$.

```{r}
p=list(r=0.05)

```

The Bellman equation can then be written as:

$$V_1(x,a) = \log \exp \Big( a \cdot x \cdot w  +  \frac{1}{1+r} E_{a'}  V_0(x,a')      \Big) $$ 


# solving the model

```{r}


```

# simulating the model

