---
title: "Homework on unobserved heterogeneity"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
editor_options: 
  chunk_output_type: console
---

```{r hw-hetero,echo=FALSE,results=FALSE}
if (file.exists("../../src/hw-hetero-sol.R")) source("../../src/hw-hetero-sol.R")
if (file.exists("../src/hw-hetero-sol.R")) source("../src/hw-hetero-sol.R")
```

In this homework, we are going to consider a simple model of unobserved heterogeneity. We will do this through the lense of unemployment duration.

# A simple model

We index individual by $i$ and we suppose that we observe their unemployment spell. Each period the unemployed individual chooses how much effort to put into searching for work. This effort affects the probability $\lambda_i$ of getting a job offer at wage $W_i$. The cost to achieve arrival probability is given by $c(\lambda_i)$. When he accecpts the job, we assume that the worker keeps the job for ever. The discount rate is $r$ and we will consider $r \rightarrow 0$.

The value of finding a job is given by $V_e = \frac{\rho W}{r}$. Where $\rho$ captures the tax system. The unemployed worker Bellman equation is then given by

$$
\begin{align}
V_u & = \max_\lambda b - c(\lambda)  + \lambda \frac{\rho W}{(1+r) r} + \frac{1-\lambda}{1+r} V_u
\end{align}
$$
<span class="label label-success">Question 1</span> show that for a fixed  $\lambda$ we can write:

$$  V_u =  \frac{1+r}{r+\lambda} \Big( b - c(\lambda) + \lambda \frac{\rho W}{(1+r) r} \Big) $$

We are going to assume that $c(\lambda) = \frac{a}{\gamma}  \lambda^{(1+\gamma)}$. 

<span class="label label-success">Question 2</span> Show that taking the first order condition on the effort decision, and taking the limit $r \rightarrow 0$, gives the following optimal choice for $\lambda$:

$$ \lim_{r \rightarrow 0} \lambda = \Big(  \frac{\rho W - b}{a}\Big)^{\frac{1}{1+\gamma}} $$

## Simulating data

We draw wages from a log-nornmal distribution and attach the $\lambda$ we derived previously.

```{r}
p = list(rho=0.9,a = 3, gamma =1.5)
p$n = 1000
data = data.table(B = 0, W = exp(rnorm(p$n) -3))

# solve for lambda
data[, Lb := ((p$rho*W - B)/(p$a))^(1/(1+p$gamma))]
data[, Lb := pmin(pmax(Lb,0),1)]

data[, D  := rexp(p$n, rate = Lb)]

ggplot(data,aes(x=W,y=Lb)) + geom_line() + theme_bw()
```

## Estimating

In this first section we coinsider the simple case where all individuals are homogenous. We are interested in estimating $a,\gamma$. 

<span class="label label-success">Question 3</span> Write down a function that computes the likelihood of a given data $D,W$. Evaluate your likelihood for a grid on $a,\gamma$ around the truth and show that the max is around the true value.

```{r}
lik.homo(data$W,data$D,p)
```

We note that the MLE estimator of $a$ for a given value of $\gamma$ admits a close form solution.

<span class="label label-success">Question 4</span> Derive the FOC condition for $a$ from the likelihood expression, and find a close form solution for the solution. 


# Random Effect

In this section we want to allow for individuals to have different values of $a$. We assume that the population is devided in $K$ groups, each having their own $a_k$ parameter. We are going to only use $K=3$.

Simulating data is as simple as before:

```{r}
p = list(rho=0.9,a = c(1,3,5), gamma =1.5, pk=c(0.2,0.5,0.3))

p$n = 1000
data = data.table(B = 0, W = exp(rnorm(p$n) -2))

# draw the latent type
data[, k := sample.int(3,.N,prob = p$pk,replace=T)]

# solve for lambda
data[, Lb := ((p$rho*W - B)/(p$a[k]))^(1/(1+p$gamma))]
data[, Lb := pmin(pmax(Lb,0),1)] # bound it between 0 and 1

data[, D := rexp(p$n, rate = Lb)]
```

<span class="label label-success">Question 5</span> write down a function that computes the liklihood of the data given our random effect model. Show that this likelihood picks close to the true value of $\gamma,a_1,a_2,a_3,p_1,p_2,p_3$ by varying each parameter separatly and reporting the likelihood (the $p$ parameters need to sum to 1).


Here, I would like for you to write the EM algorithm to recover the different types. We are going to write an Em to recover the $a_k$ for a given $\gamma$. Start with the true value of $gamma$ for now. You need write code that:

 1. E-step: compute the posterior probability for each individual to be of type i
 2. M-step: code that computes the $a_k$ parameters. You can use the close form you derived at the begining to do that. 

<span class="label label-success">Question 6</span> write down a function performs the EM estimation for a fixed value of $\gamma$.

Within each EM step, the likelihood has to be going up, otherwise it means that you have a an error.

<span class="label label-success">Question 7</span> wrap your EM algorithm in a grid over $\gamma$. You should be nicely picking at the true $\gamma$.

# Fixed Effect

We could also have tried a fixed effect approach, where we directly use the duration as a measurement of the true lambda. In the Fixed Effect world, we estimate an $a_i$ per individual. For a given individual you only have one $D_i,W_i$. Still however, you can write the MLE estimate of $a_i$ using the same formula that you derived at the begining of the homework.

We can try to maximize the overall likelihood using FE parameters $a_i$ by applying the following procedure:

 1. for a given value of $\gamma$ compute the MLE estimate of $a_i$ 
 2. evaluate the likelihood at the value of $\gamma$
 3. do this for a grid of $\gamma$ and report what seems to be the highest value.
 
<span class="label label-success">Question 8</span> Implement this in a function, report the estimate of $\gamma$ that it returns (please actually report a curve over $\gamma$). Use the following code to generate data. Comment on the result.

```{r}
p = list(rho=0.9, gamma =1.5)

p$n = 1000
data = data.table(B = 0, W = exp(rnorm(p$n) -2))

# draw the latent type
data[, a := runif(p$n, min=0.5, max=2.5)]

# solve for lambda
data[, Lb := ((p$rho*W - B)/a)^(1/(1+p$gamma))]
data[, Lb := pmin(pmax(Lb,0),1)] # bound it between 0 and 1

data[, D := rexp(p$n, rate = Lb)]
```

<span class="label label-success">Question 9 (bonus) </span> Repeat the fixed-effect approach for the case that we observe two sets of wage and unemployment duration. Specifically, we generate the data by the following code.

```{r}
p = list(rho=0.9, gamma =1.5)

p$n = 1000
data = data.table(B = 0, W1 = exp(rnorm(p$n) -2), W2 = exp(rnorm(p$n) -2))

# draw the latent type
data[, a := runif(p$n, min=0.5, max=2.5)]

# solve for lambda
data[, Lb1 := ((p$rho*W1 - B)/a)^(1/(1+p$gamma))]
data[, Lb2 := ((p$rho*W2 - B)/a)^(1/(1+p$gamma))]
data[, Lb1 := pmin(pmax(Lb1,0),1)] # bound it between 0 and 1
data[, Lb2 := pmin(pmax(Lb2,0),1)] # bound it between 0 and 1

data[, D1 := rexp(p$n, rate = Lb1)]
data[, D2 := rexp(p$n, rate = Lb2)]
```

