---
title: "Homework on Inference"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
---

## Analytical questions

**Question 1**: Show that the ols model $Y_i = \beta X_i + \epsilon_i$ is identified when $E(x\epsilon)=0$ and $var(x)>0$.

**Question 2**: Derive the bias of the sample variance.

## Computer Question 

```{r,message=FALSE}
require(kableExtra)
require(ggplot2)
require(texreg)
options(knitr.table.format = "html") 
```

We are going to reproduce an exerice similar to the example for the computation of standard error.  Start by downloading the CPS data from [here](http://cameron.econ.ucdavis.edu/research/cameron_miller_JHR_files%20to%20share.zip). For this part load the 

```{r}
require(readstata13) #to install, run install.packages("readstata13")
data = read.dta13("/Users/thibautlamadon/Downloads/files to share/CPS_2012_micro.dta")
data = data.table(data)
data$age = as.numeric(data$age)
```

Next generate a fictuous policy that you randomly assigned at the state level. Run the regression and report standard errors given by R for one draw of the poilcy.

```{r,results='asis'}
data <- data[,fp := runif(1)>0.5, statefip]
fit1 = lm(lnwage ~ fp + yrseduc + age + I(age^2) + statefip ,data)
htmlreg(fit1,single.row=TRUE,omit.coef="state")
```

Now this is surprising. To gain understanding on what is happening we will use the real data, but we will also generate our own data in a where we control exactly what is happening.

### Classical DGP

Let's start by reassuring outselves. Let's use an IID dgg, run the regression and check the significance.

 - regress without using `fp` variable, then use the `predict command` to get predicted value
 - compute an "homoskedastic" s.e. for the residuals by taking the variance of the residuals
 - simulate a fake outcome by adding a truly i.i.d. normal error to your predicted value
 - run the regression including `fp` and collect its coefficient
 - run the last steps 500 times and compute the quantiles of the regression coefficient

### Heteroskedastic $\Omega$

First we want to compute heteroskedastic robust standard errors. 

### State clustered $\Omega$

### Boostrap at the state level







