---
title: "Static Labor Supply"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r laborsupply, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(testthat)
require(xtable)
require(pander)
```

You can find the source code for this file in the class repository. The direct link is [here](https://raw.githubusercontent.com/tlamadon/econ-34430/master/src/static-labor-supply.Rmd).

Let's start with studying static labor supply. We will consider the decision of the agent under the following rule:

$$
\max_{c,h} \frac{c^{1+\eta}}{1+\eta} - \beta \frac{h^{1+\gamma}}{1+\gamma}\\
\text{s.t. } c = \rho \cdot w\cdot h -r + \mu - \beta_0 \cdot 1[h>0] \\ 
$$
The individual takes his wage $w$ as given, he chooses hours of work $h$ and consumption $c$ subject to a given non labor income $\mu$ as well as a tax regime defined by $\rho,r$. $\beta_0$ is a fixed cost associated with working.

We note already that the non labor income can control for dynamic labor supply since we can have $\mu= b_t - (1+r)b_{t+1}$. This is part of a larger maximization problem where the agents choose optimaly $b_t$ over time. We will get there next time.

## Interior solution

The first order conditions give us $w(wh +r - \mu)^\eta = \beta h^\gamma$. There is no closed-form but we can very quickly find an interior solution by using Newton maximization on the function $f(x) = w(wh +r - \mu)^\eta - \beta h^\gamma$. We iterate on 

$$x \leftarrow x - f(x)/f'(x).$$

```{r}
# function which updates choice of hours using Newton step
# R here is total unearned income (including taxes when not working and all)
ff.newt <- function(x,w,R,eta,gamma,beta) {
  f0 = w*(w*x + R)^eta - beta*x^gamma
  f1 =  eta*w^2 * (w*x + R)^(eta-1) - gamma * beta *x^(gamma-1)
  x  = x - f0/f1 
  x  = ifelse(w*x + R<=0, -R/w + 0.0001,x) # make sure we do not step out of bounds for next iteration
  x  = ifelse(x<0, 0.0001,x)
  x
}
```

## Simulating data

We are going to simulate a data set where agents will choose participation as well as the number of hours if they decide to work. To do that we will solve for the interior solution under a given tax rate and compare this to the option of no-work.

```{r, results='hide'}
p  = list(eta=-1.5,gamma = 0.8,beta=1,nu=0.5) # define preferences
tx = list(rho=1,r=0) # define a simple tax
N=50000
simdata = data.table(i=1:N,X=rnorm(N))
simdata <- simdata[,lw := X     + rnorm(N)*0.2];      # add a wage which depends on X
simdata <- simdata[,mu := exp(0.3*X + rnorm(N)*0.2)]; # add non-labor income that also depends on X

# we then solve for the choice of hours and consumption
simdata <- simdata[, h := pmax(-mu+tx$r ,0)/exp(lw)+1] # starting value
# for loop for newton method (30 should be enough, it is fast)
for (i in 1:30) {
  simdata[, h := ff.newt(h,tx$rho*exp(lw),mu-tx$r,p$eta,p$gamma,p$beta) ]
}

# attach consumption, value of working
simdata <- simdata[, c  := tx$rho*exp(lw)*h + mu];
simdata <- simdata[, u1 := c^(1+p$eta)/(1+p$eta) - p$beta * h^(1+p$gamma)/(1+p$gamma) ];
```

At this point we can regress $\log(w)$ on $\log(c)$ and $\log(h)$ and find precisely the parameters of labor supply:

```{r}
pander(summary(simdata[,lm(lw ~ log(c) + log(h))]))
```


The regression still works, among ecah individual who chooses to work, the FOC is still satified.

```{r}
pander(summary(simdata[,lm(lw ~ log(c) + log(h))]))
```

## Heterogeneity in $\beta$

Finally we want to add heterogeneity in the $\beta$ parameter. 

```{r, results="hide"}
simdata <- simdata[,betai := exp(p$beta*X+rnorm(N)*0.1)]
simdata <- simdata[, h := pmax(-mu+tx$r ,0)/exp(lw)+1]
for (i in 1:30) {
  simdata <- simdata[, h := ff.newt(h,tx$rho*exp(lw),mu-tx$r,p$eta,p$gamma,betai) ]
}

# attach consumption
simdata <- simdata[, c  := exp(lw)*h + mu];

# let's check that the FOC holds
sfit = summary(simdata[,lm(lw ~ log(c) + log(h) + log(betai))])
expect_equivalent(sfit$r.squared,1)
expect_equivalent(coef(sfit)["log(c)",1],-p$eta)
expect_equivalent(coef(sfit)["log(h)",1],p$gamma)

sfit2 = summary(simdata[,lm(lw ~ log(c) + log(h))])
expect_false(coef(sfit2)["log(c)",1]==-p$eta)

sfit3 = summary(simdata[,lm(lw ~ log(c) + log(h) + X)])
```

```{r, ,results='asis'}
pander(sfit2)
```

Next we check the non linear moment condition 

$$ E \left[ \log(w) - \gamma \log h - \eta \log c + \nu x   | W,R,X \right] $$

```{r, ,results='asis'}

simdata[, u0 := lw - p$gamma * log(h) + p$eta * log(c) - p$beta * X ]

simdata[, mean(u0)]
simdata[, mean(u0 * lw)]
simdata[, mean(u0 * X)]

ff <- function(par) {
  simdata[, u0 := lw - par[1] * log(h) + par[2] * log(c) - par[3] * X ]
  m1 = simdata[, mean(u0 * mu)]
  m2 = simdata[, mean(u0 * lw)]
  m3 = simdata[, mean(u0 * X)]
  return(m1^2 + m2^2 + m3^2)
}

ff(c(p$gamma,p$eta,p$beta))

val = rep(0,100)
sup = seq(-0.2,0.2,l=100)
for (i in 1:100) {
  val[i] = ff(c(p$gamma + sup[i] ,p$eta,p$beta))
}
plot(sup,val)
for (i in 1:100) {
  val[i] = ff(c(p$gamma ,p$eta + sup[i],p$beta))
}
plot(sup,val)
for (i in 1:100) {
  val[i] = ff(c(p$gamma ,p$eta,p$beta + sup[i]))
}
plot(sup,val)

res = optim(c(p$gamma,p$eta,p$beta),ff)
res = optim(c(p$gamma + runif(1),p$eta + runif(1),p$beta + runif(1)),ff,control = list(abstol=1e-8))

```


## Simple case of $\eta=0$

```{r}
p  = list(eta=0,gamma = 0.8,beta=1,beta0=0) # define preferences
tx = list(rho=1,r=0) # define a simple tax
N=1000
simdata = data.table(i=1:N,X=rnorm(N))
simdata <- simdata[,lw := X     + rnorm(N)*0.2];      # add a wage which depends on X
simdata <- simdata[,mu := exp(0.3*X + rnorm(N)*0.2)]; # add non-labor income that also depends on X
simdata <- simdata[,eps := rnorm(N)*0.1]
simdata <- simdata[,betai := exp(0.5*X+eps)]
simdata <- simdata[, h := (tx$rho*exp(lw)/betai)^(1/p$gamma)]

sfit3 = summary(simdata[,lm(log(h) ~ lw + X)])
pander(sfit3)

sfit4 = summary(simdata[,lm(lw ~ log(h) + X)])
pander(sfit4)

```

Then we can construct a counter-factual revenue

```{r}
p2  = list(eta=0,gamma = 1/sfit3$coefficients["lw","Estimate"],beta=1,beta0=0)
tx2 = tx
tx2$rho = 0.9
simdata <- simdata[, h2 := (tx2$rho*exp(lw)/betai)^(1/p2$gamma)]

simdata[, list(totearnings =mean(exp(lw+h)), R1=mean((1-tx$rho)*exp(lw+h)),R2=mean((1-tx2$rho)*exp(lw+h2)) ,R3=mean((1-tx2$rho)*exp(lw+h)) )]


```
