---
title: "ECON 21130 - R markdown tutorial"
author: "Wooyong Lee"
output: html_document
header-includes:
- \usepackage{amsfonts,amssymb,amsmath}
---
<!-- customize the title and the author in the above code -->

This is a tutorial for R markdown. It also serves as a sample R markdown code. Please check the source code of this document.


# What is R markdown?

- Think of it as a source code for creating a webpage that can embed R code
- You write a source code and "compile" it to create a webpage


# Creating an R markdown document

In RStudio, click **File -> New File -> R Markdown...**.

RStudio will create a skeleton file with some basic explanations. Read those.

You can compile the source code into HTML or PDF. I recommend HTML because it can adjust fontsize, linebreak, etc. according to the device.


# R markdown cheatsheet

- First of all, check the R markdown cheatsheet:
https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

- Especially, 
check **3. Markdown**.

- It covers most of 
what you need to know to write plain text

<!-- as in the cheatsheet, '*' also works for the unordered list -->


# LaTeX

If you know LaTeX, a popular language for writing academic documents, you can use it in R markdown.

R markdown understands LaTeX syntax --- just write them.

<!-- if you are familiar with LaTeX and want to use LaTeX packages in R markdown, include them in the "header-includes" declaration in the header as in this document -->

For the class, we use it to write math symbols and equations, which we discuss below.


# Writing math symbols and equations

Please check the source code of this document to see how to write math equations.

### Examples - read the source code!

<!-- $$ and $$ displays a math equation in a separate line -->

Pythagorean theorem.

$$
  a^2 + b^2 = c^2
$$

<!-- $...$ creates an inline math expression -->

Arithmetic mean $\geq$ Geometric mean $\geq$ Harmonic mean.

$$
  \frac{1}{n}\sum_{i=1}^N X_i \geq \left( \prod_{i=1}^N X_i \right)^{1/N} \geq \frac{1}{\sum_{i=1}^N 1/X_i}.
$$

Greek letters, plain texts and multiple lines.

$$
  \begin{aligned}
  line 1 &= \alpha + \beta   \\
  \text{plain text} &= \gamma + \delta  \\
  line 3 &= \lambda + \sigma \\
  \textbf{bold plain text} &= \Lambda + \Sigma
  \end{aligned}
$$

Caligraphic, blackboard bold and bold.
$$
  \mathcal{F}, \mathbb{F}, \mathbf{F}.
$$

If you need a math symbol but you don't know what to type, go to http://detexify.kirelabs.org/classify.html and draw the symbol using your mouse! For example, you can draw $\pm$ there and see that the command is "pm".


# Embedding R

- The main advantage of using R markdown is that we can run R within the document.

- In other words, we can alternate between plain text and R code.

- In terms of homework submission, this is handy because you don't need to separately submit the code or copy the R output into the document.

Below is an example of the R code (which we shall call "chunk") run within the document.

```{r}
Sys.setenv(lang="EN") ### this command sets the R locale to English ###
courseNo = "ECON 21130"
print(courseNo)
```

- Note that it displays not only the code but also the result.

- Also, the R chunks are cumulative; that is, the variable `courseNo` defined in the above can be used in the later chunks. 

Below is an example.

```{r}
courseTitle = "Topics in Microeconometrics"
print(paste(courseNo, courseTitle, sep = " - "))
```

- R markdown offers options to configure how the R chunks are displayed. For example, you can decide whether to hide the code, whether to hide the result, whether to display the error message, etc.

- For a list of such options, check **5. Embed code** of the R markdown cheatsheet.

- For example, `eval=FALSE` option tells R to not execute the code.

Below is an example.

```{r, eval=FALSE}
ROOT2 = 1.414
```

In the above chunk, I specified `eval=FALSE` option. So the above chunk is not executed, which means `ROOT2` is not defined.

In the below chunk, we can confirm that `ROOT2` is not defined.

```{r, error=TRUE}
print(ROOT2)
```

If you check the source code of this document, you will notice that I specified `error=TRUE` in the above chunk. 

You must specify this option if you want R to print the error message. Otherwise the source code will not be compiled into the HTML document due to the error.






















